# Hurricanes most harmful, Hail causes the greatest economic consequences
output: 
  html_document:
    keep_md: true
    
```{r Global}
library(data.table)
library(ggplot2)
opts_chunk$set("echo"= TRUE)
```

### Questions
- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
- Across the United States, which types of events have the greatest economic consequences?


## Synopsis
describes and summarizes the data analysis in less than 10 sentences?
 
## Data Processing
describes how the data were loaded into R and processed for analysis?

### Loading and preprocessing the data

```{r GetData, cache=TRUE}
l_DestFile <- "StormData.bz2"
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              destfile = l_DestFile, mode = "wb")

StormData <- read.csv(l_DestFile)

head(StormData)

str(StormData)
```

## Analysis
Does the analysis include description and justification for any data transformations?

## Results
the main results are presented?
at least one figure, no more than three
descriptive captions (i.e. there is a description near the figure of what is happening in the figure)?

```{r fig.width=7, fig.height=6}

```
